Step by Step: Using Step Selection Functions to Simulate Dispersal and Assess Landscape Connectivity

### 2021-06-24 Chapter_1/04_Manuscript/Manuscript.tex
various updates

%------------------------------------------------------------------------------
%	Abstract
%------------------------------------------------------------------------------
\newpage
\begin{abstract}
Dispersal is an important process that allows species to avoid inbreeding, to
colonize new habitats and to reinforce non-viable subpopulations. Successful
dispersal thus represents a crucial pre-requisite for long-term species
persistence in wild animal populations. However, the ability to disperse is
contingent a sufficient degree of landscape connectivity, which is why the
estimation of connectivity and identification of dispersal corridors has become
a task of extraordinary importance for conservation authorities worldwide.

Over the past two decades, ecologists have primarily relied on analytical tools
such as least-cost analysis and circuit theory to model and investigate
landscape connectivity. Despite their usefulness for a diverse suite of
ecological applications, both methods make several restricting assumptions that
limit their suitability in reality. To address these shortcomings, dispersal
simulations from individual-based movement models have been proposed and
applied. Yet, due to the almost infinite amount of non-trivial decisions a
modeler faces when parametrizing such models, a unified and objective framework
is missing.

@@

Here, we showcase the use of integrated step selection functions to simulate
dispersal of the endangered African wild dog across the world's largest
transboundary conservation area, the Kavango-Zambezi Transfrontier Conservation
Area (KAZA-TFCA). For this, we utilize data collected on 16 dispersing wild dog
coalitions in combination with relevant habitat covariates. We analyse the data
using integrated step selection functions and parametrize a fully mechanistic
movement model rendering wild dog dispersal. Based on this model, we simulate
80'000 dispersers, originating from protected areas and moving across the extent
of the KAZA-TFCA. We then generate a heatmap indicating regions frequently
visited by dispersers, use network theory to reveal dispersal hotspots and
crucial bottlenecks across the study area. Finally, we discuss the benefits and
pitfalls of such dispersal simualtions and highlight potential improvements to
be made in the future.

@@

% Importance of Dispersal & Connectivity
\subsection{Importance of Dispersal \& Connectivity (90\%)}
Dispersal is defined as the movement of individuals away from their natal
location to the site of first reproduction \cite{Howard.1960}. It is a vital
process governing the dynamics wild animal populations that are distributed in
space \citep{Hanski.1998, Clobert.2012} and may strongly affect population
dynamics at different spatial and social scales \citep{Hanski.1999a,
Clobert.2012}. Dispersal allows species to avoid inbreeding and maintain genetic
diversity \citep{Perrin.1999, Perrin.2000, Frankham.2002, Leigh.2012,
Baguette.2013}, to rescue small and non-viable populations \citep{Brown.1977},
and to promote the colonization of unoccupied habitats \citep{Hanski.1999b,
MacArthur.2001}. However, successful dispersal requires a sufficient degree of
functional connectivity \citep{Fahrig.2003, Clobert.2012}, which is why the
identification and protection of major dispersal corridors has become an
important task in conservation science \citep{Doerr.2011, Rudnick.2012}. In
order to pinpoint relevant dispersal hotspots, reliable information on movement
behavior during dispersal and knowledge about factors that limit dispersal and
connectivity is paramount \citep{Baguette.2013, Vasudev.2015}.

% Advancements in GPS Technology & Movement Ecology
\subsection{Advancements in GPS Technology \& Movement Ecology (90\%)}
Thanks to novel technologies developed over the past decades, particularly of
GPS/Satellite radio-collars, the use of GPS data to study animal dispersal and
connectivity has accelerated \citep{Elliot.2014, Jonsson.2016, Williams.2019}.
Additionally, the advent of publicly accessible satellite imagery and
sophisticated remote sensing techniques to represent the physical landscape
through which individuals disperse have heralded a ``golden age of animal
tracking'' \citep{Kays.2015}. Concurrently, the availability of large amounts of
empirical data and an increased computational power have led to the development
of numerous techniques to study dispersal movements and highlight critical
corridors between subpopulations \citep{Boyce.2002, Fortin.2005, Cushman.2010,
Zeller.2012, Diniz.2020}.

% Resource Selection & Connectivity
\subsection{Resource Selection \& Connectivity (90\%)}
\textit{Resource selection functions} \citep{Boyce.2002} and derived methods
such as \textit{step selection functions} \citep{Fortin.2005} and \textit{path
selection functions} \citep{Cushman.2010} have proven particularly useful for
studying animal movement \citep{Fieberg.2020} and modelling connectivity
\citep{Diniz.2020}. These methods allow estimating habitat preferences of the
focal species by comparing covariates at locations visited by the animal to the
same covariates at locations available to, but not visited by the animal
\citep{Boyce.2002, Fortin.2005, Cushman.2010, Thurfjell.2014}. The so estimated
preferences can then be used to predict a permeability surface, indicating the
expected ease at which an animal can traverse a given area \citep{Spear.2010,
Zeller.2012, Etherington.2016}. Ultimately, the permeability surface serves as
input to a connectivity model that is used to reveal movement corridors
\citep{Diniz.2020}. In this regard, two of the most prominent connectivity
models are least-cost path analysis (LCP analysis; \citealp{Adriaensen.2003})
and circuit theory (CT \citealp{McRae.2006, McRae.2008}), both graph-based
methods that estimate conductance of the landscape. Despite their intuitive
nature and ease of use, both methods make rigorous assumptions about animal
movement that are often not fulfilled in reality \citep{Diniz.2020}.

% Issues with Least-Cost Paths & Circuit Theory
\subsection{Issues with Least-Cost Paths \& Circuit Theory (90\%)}
In LCP analysis, for instance, a least costly path always exists, even if
associated movement costs are unreasonably high and will never be incurred by a
dispersing individual. The method also presumes that animals have an infinite
perceptual range, a preconceived end-point in mind, and choose a cost-minimizing
route accordingly. These assumptions may be reasonable for migrating animals,
yet they are unlikely to hold for dispersers, which typically move over long
distances into unknown territory \citep{Koen.2014, Abrahms.2017, Cozzi.2020}.
Finally, LCPs are only one pixel wide, meaning that their absolute size depends
on the resolution of chosen covariate layers \citep{Diniz.2020}. Although some
of these deficiencies can be addressed using less stringent versions of the LCP
algorithm (e.g. least-cost \textit{corridors} \citep{Pinto.2009},
\textit{thresholded} least-cost paths \citep{Landguth.2012}, and
\textit{randomized} least-cost paths \citep{Panzacchi.2016, VanMoorter.2021}), a
certain degree of arbitrariness in the assumptions remains. CT entails similarly
unreasonable restrictions. Because CT only considers movements from the source
cell to its 4 or 8 adjacent cells, it implicitly posits a perceptual range of a
single pixel. As such, the perceptual range cannot be defined based on
biological observations, but is imposed by the resolution of the reference grid.
CT is also built on the assumption of a complete random walk \citep{Diniz.2020},
implying that directional biases cannot be rendered, albeit being very common in
dispersal movements \citep{Cozzi.2020, Hofmann.2021}. Ultimately, neither LCP
analysis nor CT are capable of rendering the temporal dimension of dispersal.
Statements about the expected duration required to traverse a certain corridor
are therefore impossible, albeit being important factors. Likewise, because
movement is not modelled explicitely, neither of the methods allows to render
interactions between movement and habitat preferences of the focal species. This
implies that connectivity mainly arises as a result of the landscape structure
and is referred to as structural connectivity. Structural connectivity stands in
contrast to functional connectivity, which also renders the behavioral response
of the animal with respect to prevailing habitat conditions
\citep{Tischendorf.2000}. Even though functional connectivity is more difficult
to estimate, a functional view is the ultimate goal in conservation science
because it has direct consequences for gene flow \citep{Baguette.2013}.

% What about IBMMs?
\subsection{What about IBMMs? (90\%)}
To address the issues inherent to LCPs and CT, individual-based movement models
(IBMMs) have been proposed and applied \citep{Diniz.2020}. In these models,
dispersal is simulated explicitly, based on movement rules that determine how
individuals move over and interact with the prevailing landscape
\citep{Kanagaraj.2013, Clark.2015, Allen.2016, Hauenstein.2019, Zeller.2020,
Vasudev.2021}. Using the simulated trajectories, one can calculate a set of
connectivity metrics, such as interpatch-connectivity and traversal frequency
across the landscape to reveal major dispersal hotspots \citep{Kanagaraj.2013,
BastilleRousseau.2018, Hauenstein.2019, Zeller.2020}. Even though IBMMS can be
employed to overcome any of the shortcomings intrinsic to LCPs and CT, as well
as to provide a more functional view on connectivity, they can be challenging to
fit and require vast amounts of data collected during dispersal
\citep{Diniz.2020}.

@@

% Study Species & Study Area
\subsection{Study Species \& Study Area (90\%)}
One of the species for which long-term viability relies on sufficient landscape
connectivity is the endangered African wild dog \textit{Lycon pictus}. While
once present across entire sub-Saharan Africa, wild dogs have disappeared from a
vast majority of their historic range due to persecution by humans, habitat
fragmentation and destruction, and deadly diseases \citep{Woodroffe.2012,
Woodroffe.2020}. As of today, only 6'000 free-ranging individuals remain in
small and spatially scattered subpopulations \citep{Woodroffe.2012}. Within
those subpopulations, wild dogs form cohesive packs comprising 8 to 12 adults
and their offspring \cite{McNutt.1995}. After reaching sexual maturity, male and
female offspring form same-sex coalitions and disperse from their natal pack
\citep{McNutt.1996, Behr.2020}. New packs are formed when dispersing coalitions
join unrelated opposite-sex dispersing coalitions \citep{McNutt.1996}.
Dispersing wild dogs can cover several hundred kilometers across a variety of
landscapes \citep{DaviesMostert.2012, Masenga.2016, Cozzi.2020, Hofmann.2021}.
One of the few strongholds for this species lies near the Moremi Game Reserve in
northern Botswana, which is part of the world's largest transboundary protected
area, namely the Kavango-Zambezi Transfrontier Conservation Area (KAZA-TFCA).
This area has originally been intended to facilitate migration of elephants, but
is expected to benefit a multitude of other species \citep{Elliot.2014,
Brennan.2020, Hofmann.2021}.

% Previous Paper
\subsection{Previous Paper (90\%)}
In a previous study, we assessed landscape connectivity for dispersing African
wild dogs within the KAZA-TFCA using a least-cost corridor approach
\citep{Hofmann.2021}. For this, we fitted a basic habitat selection model based
on which we predicted landscape permeability. We now expand on this knowledge
and use ISSF to develop a more detailed movement model of dispersing wild dogs
(\Cref{GraphicalAbstract}). We then use this model to simulate dispersers moving
across the KAZA-TFCA. Based on simulations, we compute heatmaps and identify
potential dispersal hotspots. We also showcase how network metrics relevant to
landscape connectivity can be computed. Our results show that a simulation-based
approach yields several major benefits over traditional connectivity modeling
techniques. Most importantly, simulations provide a more generic view on how
connectivity emerges and to which degree connectivity depends on the dispersal
duration. In addtion, by generating proper dispersal trajectories, network
theory can be applied to calculate network metrics that are pertinent to
connectivity analysis.

@@

\subsection{Movement Model (80\%)}
We used ISSFs to parametrize a mechanistic movement model for dispersing wild
dogs \citep{Avgar.2016}. To conduct ISSF analysis, we paired each realized step
with 24 random steps. An observed and its 24 random steps formed a stratum and
received a unique identifier. We generated random steps by sampling random
turning angles from a uniform distribution (\(-\pi, +\pi\)) and step lengths
from a gamma distirbution that was fitted to realized steps (scale = 6'308,
shape = 0.37). Along each step, we extracted and averaged spatial covariates
using the {\tt velox} package \citep{Hunziker.2017}. We also calculated the
movement metrics \textsf{sl}, \textsf{log(sl)}, and \textsf{cos(ta)} for each
observed and random step. To facilitate model convergence, we standardized all
continuous covariates to a mean of zero and a standard deviation of one. Since
correlation among covariates was low (\(|r| > 0.6\); \citealp{Latham.2011}), we
retained all of them for modeling.

@@

\subsection{Heatmap (100\%)}
To identify dispersal hotspots across our study area, we created a heatmap
indicating the absolute frequency at which each raster-cell in the study area
was visited by virtual dispersers. For this, we rasterized all simulated
trajectories and tallied them into a single map. If the same trajectory crossed
a raster-cell twice, it was only counted once. This way, we mitigated potential
biases resulting from trapped individuals that were moving in circles. To
achieve high performance rasterization, we used the R-package {\tt terra}
\citep{Hijmans.2020}.

\subsection{Betweenness (80\%)}
To pinpoint areas of exceptional relevance for connecting remote regions in our
study area, we converted simulated trajectories into a network and calculated
betweenness scores. For this, we overlaid the study area (including the buffer)
with a regular raster resolved at 5 x 5 km. The centerpoint of each raster-cell
acted as node in the final network and we used the simulated trajectories to
determine all transitions occurring from one raster-cell to another, as well as
the frequency at which those transitions occurred. This resulted in an edge-list
that we translated into a weighted network using the r-package {\tt igraph}
\citep{Gabor.2006}. Because {\tt igraph} handles edge weights (\(\omega\)) as
costs, we inverted the traversal frequency in each cell by applying \(\omega =
\frac{\sum_i^n{Traversal Frequency_i}/n}{Traversal Frequency_i}\). Finally, we
used the weighted network to calculate the betweenness score of each
raster-cell. Betweenness measures how often a specific raster-cell lies on a
shortest path between two other raster-cells and is therefore a useful metric to
detect movement corridors \citep{BastilleRousseau.2018}.

@@

\section{Results}
\subsection{Movement Model (60\%)}
Compared to the base model reported in \citep{Hofmann.2021}, our most
parsimonious movement model contained several additional interactions between
habitat covariates and movement covariates (\Cref{MovementModel} and
\Cref{MovementModelNumbers}). Although multiple models received an AIC weight
above zero (Table 1 in Appendix S1), we only considered results from the most
parsimonious model. Since all models with positive AIC weight contained similar
covariates (Table S1), this decision only marginally influenced subsequent
analyses. Plots that aid with the interpretation of the final model are provided
in Appendix S2.

When looking at the habitat kernel (and holding constant the movement kernel),
we find that dispersing wild dogs avoid water but prefer its proximity.
Dispersers also avoid woodlands, yet prefer shrublands or grasslands. Finally,
dispersers avoid moving through landscapes that are influenced by humans. These
results align with our previous findings reported in \cite{Hofmann.2021}.

Results from the movement kernel indicate several significant estimates.
However, except for the interaction \textsf{sl:LowActivity}, effect sizes are
relatively small, indicating that our proposal distributions for step lengths
and turning angles were only marginally biased by habitat selection. For
instance, the positive and significant estimate for \textsf{cos(ta)} indicates
that realized turning angles were slightly more directional than the turning
angles proposed by our uniform distribution. This shows that realized steps in
reality followed a von Mises Distribution with positive concentration parameter.
On the other hand, the significant and negative interaction for
\textsf{sl:LowActivity} reveals that wild dogs realized shorter steps during low
wild dog activity compared to the steps suggested by our gamma distribution.

When looking at the interactions between the movement and habitat kernel, we
observe several significant interactions, highlighting that movement behavior
differs depending on habitat conditions. For example, there's ample evidence
that the propensity to realize a log step in areas covered by water is
substantially smaller compared to the propensity of realizing a small step.
Similarly, long steps are less likely to be realized in areas covered by trees,
compared to more open regions. Conversely, it appears that steps tend to longer
be in the vicinity to water, even though the size of this effect is negligible.
Finally, it seems that dispersers move more tortuous in areas influenced by
humans, but more directional when far from water.

@@

\begin{figure}
  \begin{center}
    \includegraphics[width=\textwidth]{99_MovementModel}
    \caption{(a) Most parsimonious movement model for dispersing wild dogs. The
    model includes a habitat kernel, a movement kernel, as well as their
    interactions. The orange and blue line segments delineate the 90\%, 95\%,
    and 99\% Confidence-Intervals around the respective \(\beta\) coefficients.
    Significance codes: * \(p < 0.10\), ** \(p < 0.05\), *** \(p < 0.01\). (b)
    Results from the k-fold cross validation procedure. The upper plot shows
    rank frequencies of predicted realized scores according to model predictions
    with known preferences, whereas the lower plot shows rank frequencies when
    assuming random preferences. The blue ribbon shows the prediction interval
    around a loess smoothing regression that we fitted to ease the
    interpretation of the plots.}
    \label{MovementModel}
  \end{center}
\end{figure}

>>>>

%------------------------------------------------------------------------------
%	Abstract
%------------------------------------------------------------------------------
\newpage
\begin{abstract}
Dispersal is an important process that allows species to avoid inbreeding, to
colonize new habitats and to reinforce non-viable subpopulations. Successful
dispersal thus represents a crucial pre-requisite for long-term species
persistence in wild animal populations. However, the ability to disperse is
contingent a sufficient degree of landscape connectivity, which is why the
estimation of connectivity and identification and preservation of dispersal
corridors has become a task of extraordinary importance for conservation
authorities worldwide.

Over the past two decades, ecologists have primarily relied on analytical tools
such as least-cost analysis and circuit theory to model and investigate
landscape connectivity. Despite their usefulness for a diverse suite of
ecological applications, both methods make several restricting assumptions that
limit their suitability in reality. To address these shortcomings, dispersal
simulations from individual-based movement models have been proposed and
applied. Yet, due to the almost infinite number of non-trivial decisions a
modeler faces when parametrizing such models, a unified and objective framework
is missing.

@@

Here, we showcase the use of integrated step selection functions as a simple
individual-based and spatially explicit model to simulate dispersal of the
endangered African wild dog across the world's largest transboundary
conservation area, the Kavango-Zambezi Transfrontier Conservation Area
(KAZA-TFCA). For this, we utilize data collected on 16 dispersing wild dog
coalitions in combination with relevant habitat covariates. We analyse the data
using integrated step selection functions and parametrize a fully mechanistic
model describing how dispersing wild dogs move through the landscape. Based on
this model, we simulate 80'000 dispersers moving across the extent of the
KAZA-TFCA. We then generate a set of maps, each focused on a different aspect of
connectivity within the KAZA-TFCA. Finally, we discuss the benefits and pitfalls
of such a simulation-based approach and highlight potential improvements to be
made in the future.

@@

% Importance of Dispersal & Connectivity
\subsection{Importance of Dispersal \& Connectivity (90\%)}
Dispersal is defined as the movement of individuals away from their natal
location to the site of first reproduction \cite{Howard.1960}. It is a vital
process governing the social structure of wild animal populations that are
distributed in space \citep{Hanski.1998, Clobert.2012} and may strongly affect
population dynamics at different spatial and social scales \citep{Hanski.1999a,
Clobert.2012}. Dispersal allows species to avoid inbreeding and maintain genetic
diversity \citep{Perrin.1999, Perrin.2000, Frankham.2002, Leigh.2012,
Baguette.2013}, to rescue small, non-viable populations \citep{Brown.1977}, and
to promote the colonization of unoccupied habitats \citep{Hanski.1999b,
MacArthur.2001}. However, successful dispersal requires a sufficient degree of
landscape connectivity \citep{Fahrig.2003, Clobert.2012}, which is why the
identification and protection of major dispersal corridors has become a
fundamental task in conservation science \citep{Nathan.2008, Doerr.2011,
Rudnick.2012}. Importantly, the ability to pinpoint relevant dispersal hotspots
requires information on movement behavior during dispersal and knowledge about
factors that limit dispersal and therefore connectivity \citep{Baguette.2013,
Vasudev.2015}.

% Advancements in GPS Technology & Movement Ecology
\subsection{Advancements in GPS Technology \& Movement Ecology (90\%)}
Thanks to novel technologies developed over the past decades, particularly of
GPS/Satellite radio-collars, the use of GPS data to study animal dispersal and
connectivity has accelerated \citep{Elliot.2014, Jonsson.2016, Williams.2019}.
Additionally, the advent of publicly accessible satellite imagery and
sophisticated remote sensing techniques to represent the physical landscape
through which individuals disperse have heralded a ``golden age of animal
tracking'' \citep{Kays.2015}. Concurrently, the availability of large amounts of
empirical data and an increased computational power have led to the development
of numerous techniques to study dispersal and highlight critical corridors
between subpopulations \citep{Boyce.2002, Fortin.2005, Cushman.2010,
Zeller.2012, Diniz.2020}.

% Resource Selection & Connectivity
\subsection{Resource Selection \& Connectivity (90\%)}
\textit{Resource selection functions} \citep{Boyce.2002} and derived methods
such as \textit{step selection functions} \citep{Fortin.2005} and \textit{path
selection functions} \citep{Cushman.2010} have proven particularly useful for
studying animal movement \citep{Fieberg.2020} and modeling connectivity
\citep{Diniz.2020}. These methods allow estimating habitat preferences of the
focal species by comparing covariates at locations visited by the animal to the
same covariates at locations available to, but not visited by the animal
\citep{Boyce.2002, Fortin.2005, Cushman.2010, Thurfjell.2014}. The so estimated
preferences can then be used to predict a permeability surface, indicating the
expected ease at which an animal can traverse a given area \citep{Spear.2010,
Zeller.2012, Etherington.2016}. Ultimately, the permeability surface serves as
input to a connectivity model that is used to reveal movement corridors
\citep{Diniz.2020}. In this regard, two of the most prominent connectivity
models are least-cost path analysis (LCP analysis; \citealp{Adriaensen.2003})
and circuit theory (CT \citealp{McRae.2006, McRae.2008}), both graph-based
methods that estimate conductance of the landscape. Despite their intuitive
nature and ease of use, both methods make rigorous assumptions about animal
movement that are often not fulfilled in reality \citep{Diniz.2020}.

% Issues with Least-Cost Paths & Circuit Theory
\subsection{Issues with Least-Cost Paths (90\%)}
In LCP analysis, for instance, a least costly path always exists, even if
associated movement costs are unreasonably high and will never be incurred by a
dispersing individual. The method also presumes that animals have an infinite
perceptual range and a preconceived end-point in mind, such that they choose a
cost-minimizing route accordingly. These assumptions may be reasonable for
migrating animals, yet they are unlikely to hold true for dispersers, which
typically move over long distances into unknown territory \citep{Koen.2014,
Abrahms.2017, Cozzi.2020}. Another issue of LCPs analysis concerns the fact that
least-costly routes, by their very nature, are only one pixel wide. This implies
that the width of inferred movement routes depends on the resolution of chosen
covariate layers and may not be biologically meaningful \citep{Diniz.2020}.
Although some of these deficiencies can be addressed using less stringent
versions of the LCP algorithm (e.g. least-cost \textit{corridors}
\citep{Pinto.2009}, \textit{thresholded} least-cost paths \citep{Landguth.2012},
and \textit{randomized} least-cost paths \citep{Panzacchi.2016,
VanMoorter.2021}), a certain degree of arbitrariness remains.

\subsection{Issues with Circuit Theory (90\%)}
CT entails similarly unreasonable restrictions that are hardly ever met. Because
CT only allows movements from a source cell to its 4 or 8 adjacent cells, it
implicitly posits a perceptual range of a single pixel. Assuming that covariate
layers are usually resolve between 30 m and 1 km, this hardly ever renders the
true capability of animals to perceive the environment. Moreover, CT builds on
the assumption of a complete random walk \citep{Diniz.2020}, implying that
directional biases cannot be rendered. Nevertheless, directionality is a
commonly observed characteristic in animal movement \citep{Bovet.1991,
Schultz.2001}, especially in dispersing individuals \citep{Cozzi.2020,
Hofmann.2021}.

\subsection{Issues of Both Methods}
Finally, neither LCP analysis nor CT are capable of rendering the temporal
dimension of dispersal. Statements about the expected duration required to
traverse a certain corridor are therefore impossible. Likewise, because movement
is not modeled explicitly, neither of the methods allows to render interactions
between movement and habitat preferences of the focal species. Connectivity
therefore merely arises in result to the landscape structure, which is usually
referred to as structural connectivity. While structural connectivity yields
insights in the \textit{potential} of the landscape to be traversed, it does not
enable to quantify the \textit{actual} gene flow through the area. Consequently,
a functional on connectivity, which also renders the behavioral response of the
animal with respect to prevailing habitat conditions, is often more desirable
\citep{Tischendorf.2000, Baguette.2013}.

% What about IBMMs?
\subsection{What about IBMMs? (90\%)}
To address the issues inherent to LCPs and CT, individual-based movement models
(IBMMs) have been proposed and applied \citep{Diniz.2020}. In these models,
dispersal is simulated spatially explicitly, based on movement rules that
determine how individuals move over and interact with the prevailing landscape
\citep{Gustafson.1996, Gardner.2004, Graf.2007, KramerSchadt.2004, Revilla.2004,
Revilla.2008, Kanagaraj.2013, Clark.2015, Allen.2016, Hauenstein.2019,
Zeller.2020, Vasudev.2021}. Using the simulated trajectories, one can calculate
a set of connectivity metrics, such as inter-patch connectivity and traversal
frequency to reveal major dispersal corridors \citep{Kanagaraj.2013,
BastilleRousseau.2018, Hauenstein.2019, Zeller.2020}. However, while IBMMS can
be employed to overcome any of the shortcomings intrinsic to LCPs and CT and
provide a more functional view on connectivity, they can be challenging to fit
and require vast amounts of data collected during dispersal \citep{Diniz.2020}.
Moreover, most individual-based models are heavily influenced by subjective
decisions, rendering replication and validation of results difficult.

@@

% Study Species & Study Area
\subsection{Study Species \& Study Area (90\%)}
One of the species for which long-term viability relies on sufficient landscape
connectivity is the endangered African wild dog \textit{Lycon pictus}. While
once present across entire sub-Saharan Africa, wild dogs have disappeared from a
vast majority of their historic range due to persecution by humans, habitat
fragmentation and destruction, and deadly diseases \citep{Woodroffe.2012}. As of
today, only 6'000 free-ranging individuals remain in small and spatially
scattered subpopulations \citep{Woodroffe.2012}. Within those subpopulations,
wild dogs form cohesive packs comprising 8 to 12 adults and their offspring
\cite{McNutt.1995}. After reaching sexual maturity, male and female offspring
form same-sex coalitions and disperse from their natal pack \citep{McNutt.1996,
Behr.2020}. New packs are formed when dispersing coalitions join unrelated
opposite-sex dispersing coalitions \citep{McNutt.1996}. Dispersing wild dogs can
cover several hundred kilometers across a variety of landscapes
\citep{DaviesMostert.2012, Masenga.2016, Cozzi.2020, Hofmann.2021}. One of the
few strongholds for this species lies near the Moremi Game Reserve in northern
Botswana, which is part of the world's largest transboundary protected area,
namely the Kavango-Zambezi Transfrontier Conservation Area (KAZA-TFCA). This
area has originally been intended to facilitate migration of elephants, but is
expected to benefit a multitude of other species \citep{Elliot.2014,
Brennan.2020, Hofmann.2021}.

% Previous Paper
\subsection{Previous Paper (90\%)}
In a previous study, we assessed landscape connectivity for dispersing African
wild dogs within the KAZA-TFCA using a least-cost methods \citep{Hofmann.2021}.
Specifically, we fitted a basic habitat selection model using which we predicted
a permeability surface and computed least-cost paths and corridors. We now
expand this model and use ISSFs to develop a more detailed movement model of
dispersing wild dogs (\Cref{GraphicalAbstract}). We then employ the model to
simulate dispersers moving across the KAZA-TFCA. Based on simulations, we
generate three distinc connectivity maps, all sheding light onto a different
aspect of connectivity. With this work we hope to show that a simulation-based
approach yields several benefits over traditional connectivity modeling
techniques such as least-cost analyses and circuit theory. Most importantly,
simulations provide a more generic view on how connectivity emerges and to which
degree connectivity depends on the dispersal duration. In addtion, by generating
proper dispersal trajectories, network theory can be applied to calculate
network metrics that are pertinent to connectivity analysis.

@@

\subsection{Movement Model (80\%)}
We used ISSFs to parametrize a mechanistic movement model for dispersing wild
dogs \citep{Avgar.2016}. To conduct ISSF analysis, we paired each realized step
with 24 random steps. An observed and its 24 random steps formed a stratum and
received a unique identifier. We generated random steps by sampling random
turning angles from a uniform distribution (\(-\pi, +\pi\)) and step lengths
from a gamma distirbution that was fitted to realized steps (scale = 6'308,
shape = 0.37). Along each step, we extracted and averaged spatial covariates
using the {\tt velox} package \citep{Hunziker.2021}. We also calculated the
movement metrics \textsf{sl}, \textsf{log(sl)}, and \textsf{cos(ta)} for each
observed and random step. To facilitate model convergence, we standardized all
continuous covariates to a mean of zero and a standard deviation of one. Since
correlation among covariates was low (\(|r| > 0.6\); \citealp{Latham.2011}), we
retained all of them for modeling.

@@

\subsection{Convegence (80\%)}
To verify that the number of simulated individuals sufficed to ensure reliable
estimates of connectivity, we evaluated how the relative traversal frequency
across the landscape depended on the number of simulated trajectories.
Specifically, we placed 1'000 ``checkpoints'' each with an extent of 5 km x 5 km
at random locations inside the main study area. We then determined the relative
traversal frequency by simulated trajectories through each checkpoint for
different numbers of simulations (1 to 50'000). We replicated each random sample
100 times and computed the mean across replicates, as well as the 95\%
xx-interval. Note that we only considered the 50'000 trajectories originating
from the main study area.

\subsection{Heatmap (100\%)}
To identify dispersal hotspots across our study area, we created a heatmap
indicating the absolute frequency at which each raster-cell in the study area
was visited by virtual dispersers \citep{Hauenstein.2019, Peer.2008}. For this,
we rasterized all simulated trajectories and tallied them into a single map. If
the same trajectory crossed a raster-cell twice, it was only counted once,
thereby mitigating potential biases resulting from individuals that were trapped
and moved in circles. To achieve high performance rasterization, we used the
R-package {\tt terra} \citep{Hijmans.2020}.

\subsection{Betweenness (80\%)}
To pinpoint areas of exceptional relevance for connecting remote regions inside
our study area, we converted simulated trajectories into a network and
calculated betweenness scores \citep{BastilleRousseau.2018}. For this, we
overlaid the study area (including the buffer) with a regular raster resolved at
5 x 5 km. The centerpoint of each raster-cell acted as node in the final network
and we used the simulated trajectories to determine all transitions occurring
from one raster-cell to another, as well as the frequency at which those
transitions occurred. This resulted in an edge-list that we translated into a
weighted network using the r-package {\tt igraph} \citep{Gabor.2006}. Because
{\tt igraph} handles edge weights (\(\omega\)) as costs, we inverted the
traversal frequency in each cell by applying \(\omega = \frac{\sum_i^n{Traversal
Frequency_i}/n}{Traversal Frequency_i}\). Finally, we used the weighted network
to calculate the betweenness score of each raster-cell. Betweenness measures how
often a specific raster-cell lies on a shortest path between two other
raster-cells and is therefore a useful metric to detect movement corridors
\citep{BastilleRousseau.2018}.

@@

\section{Results}
\subsection{Movement Model (80\%)}
Compared to the base model reported in \citep{Hofmann.2021}, our most
parsimonious movement model retained several additional interactions between
habitat covariates and movement covariates (\Cref{MovementModel} and
\Cref{MovementModelNumbers}). Although several models received an AIC weight
above zero (Table 1 in Appendix S1), we only considered results from the most
parsimonious model for simplicity. All models with positive AIC weight included
similar covariates (Table S1), so this decision only marginally influenced
subsequent analyses. Plots that aid with the interpretation of the final model
are provided in Appendix S2.

Assuming that all other covariates are held constant, the habitat kernel reveals
that dispersing wild dogs avoid water but prefer its proximity. Similarly,
dispersers avoid areas that are covered by woodlands, yet prefer regions covered
by shrublands or grasslands. Finally, dispersers avoid movement through
landscapes that are dominated by humans. Effect sizes are strong and, except for
effect for \textsf{distance to water}, statistically clear on the 5\%
significance level.

With regards to the movement kernel, the positive estimate for \textsf{cos(ta)}
indicates that dispersers move with directional persistence, unlike what was
proposed by the uniform turning angle distribution. Moreover, directionality is
particularly pronounced when dispersers realize large steps (move quickly), as
indicated by the positive estimates for \textsf{cos(ta):sl} and
\textsf{cos(ta):log(sl)}. Finally, the negative estimate for the interaction
\textsf{sl:LowActivity} reveals that wild dogs realize shorter steps (move
slowly) outside the main activity periods during sunrise and sunset. Aside from
the interaction \textsf{sl:LowActivity}, which appears to strongly influence
movmement behavior, effect sizes are moderate, but mostly significant on the 5\%
significance level.

When looking at the interactions between the movement and habitat kernels, we
observe that movement behavior is contingent on habitat conditions. For example,
there's strong evidence that dispersers realize smaller steps in areas covered
by water or areas covered by wooldand. Conversely, it appears that steps tend to
larger in regions dominated by shrubs/grassland, and shorter when the distance
to water is high. Correspondingly, the model suggests that directionality is
lower in areas dominated by humans, yet more pronounced when dispersers are far
from water. However, except for the effect of \textsf{sl:Water}, effect sizes
are small.

@@

\begin{figure}
  \begin{center}
    \includegraphics[width=\textwidth]{99_MovementModel}
    \caption{(a) Most parsimonious movement model for dispersing wild dogs. The
    model includes a habitat kernel, a movement kernel, as well as their
    interactions. The horizontal line segments delineate the 90\%, 95\%, and
    99\% Confidence-Intervals for the respective \(\beta\)-coefficients.
    Significance codes: * \(p < 0.10\), ** \(p < 0.05\), *** \(p < 0.01\). (b)
    Results from the k-fold cross validation procedure. The upper plot shows
    rank frequencies of realized steps according to model predictions with known
    preferences, whereas the lower plot shows rank frequencies of realized steps
    when assuming random preferences. The blue ribbon shows the prediction
    interval around a loess smoothing regression that we fitted to ease the
    interpretation of the plots. The significant correlation between rank and
    associated frequency in (b1) highlights that the most parsimonious model
    successfully outperforms a random guess (b2) and assigns comparably high
    selection scores to realized steps.}
    \label{MovementModel}
  \end{center}
\end{figure}
